---
title: "Reproducible Research Project 1"
author: "Antoinette Mills"
date: "March 6, 2016"
output: html_document
---

Requirement 1 - Code for reading in the data set and/or processing the data:
Show any code that is needed to

    Load the data (i.e. read.csv())
    Process/transform the data (if necessary) into a format suitable for your analysis

````{r}
# Load the data file
MyData <- read.csv("activity.csv", as.is = TRUE)

# Remove the NA values and store in created subset object
GoodActivityData <- MyData[complete.cases(MyData), ]

````

Requirement 2 - What is mean total number of steps taken per day?

Step 1:  Calculate the total number of steps taken per day
````{r}
#Calculate total number of steps taken per day
TotalStepsPerDay <- aggregate(steps ~ date, GoodActivityData, sum)
````

Step2: Make a histogram of the total number of steps taken each day
````{r,echo=FALSE}
hist(TotalStepsPerDay$steps, main = "Total Number Of Steps Per Day (Histogram)", xlab = "Steps per day", col="purple")
````

Step3:  Calculate and report the mean and median of the total number of steps taken per day

````{r}
#Calculate mean
MeanOfSteps<-mean(TotalStepsPerDay$steps)
print(paste("The Mean of the total number of steps per day is ",MeanOfSteps))

#Calculate median
MedianOfSteps<-median(TotalStepsPerDay$steps)
print(paste("The Median of the total number of steps per day is ",MedianOfSteps))
````

Requirement 3 - What is the average daily activity pattern?

Step 1:  Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
````{r,echo=FALSE}
#average number of steps across all days
AvgStepsAllDays <- aggregate(steps ~ interval, GoodActivityData, mean)
Intervals <- aggregate(steps ~ date, GoodActivityData, mean)

# Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
plot(AvgStepsAllDays$interval, AvgStepsAllDays$steps, type='l', main="Average Number of Steps (Interval)", xlab="Intervals", ylab="Avg # Of Steps", col="blue")
abline(h=mean(Intervals$steps, na.rm=TRUE),col="red",lwd=2)
````

Step2:  Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
````{r,echo=FALSE}
#Identify the interval index which has the highest average steps and the specific interval and average steps for that interval
print (AvgStepsAllDays[which.max(AvgStepsAllDays$steps),])
````
Interval at 835 (8:35) with a value of 206.2 steps


Requirement 4 - Imputing missing values

Step 1:  Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
````{r}
# Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
MissingValue <- MyData[!complete.cases(MyData), ]
print (paste("The number of rows with missing values is ",nrow(MissingValue)))
````

Step 2 & 3: 
````{r}

for (i in 1:nrow(MyData)) {
        if(is.na(MyData$steps[i])) {
                val <- AvgStepsAllDays$steps[which(AvgStepsAllDays$interval == MyData$interval[i])]
                MyData$steps[i] <- val 
        }
}
````


Step 4:  Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
````{r}
#Calculate total number of steps taken per day with imputed Data
TotalStepsPerDay_imputed <- aggregate(steps ~ date, MyData, sum)

#Histogram of number of steps per day with imputed Data
hist(TotalStepsPerDay_imputed$steps, main = "Total Number Of Steps Per Day (Histogram - Imputed Data)", xlab = "Steps per day", col="blue")

#Calculate mean of imputed data
MeanOfSteps_imputed<-mean(TotalStepsPerDay_imputed$steps)
print(paste("The Mean of the total number of steps per day is ",MeanOfSteps_imputed))

#Calculate median of imputed data
MedianOfSteps_imputed<-median(TotalStepsPerDay_imputed$steps)
print(paste("The Median of the total number of steps per day is ",MedianOfSteps_imputed))
````

Both the Median and the Mean have not significantly changed.  Now let's examine impact on estimates of total daily number of steps:

````{r}
#What is the impact of imputing missing data on the estimates of the total daily number of steps?
MyDifference<-sum(MyData$steps)-sum(GoodActivityData$steps)
MyDifference
````
There was a significant impact on the estimation of total steps per day (increase).

Requirement 5:  Are there differences in activity patterns between weekdays and weekends?

Step 1:  Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

````{r}
WeekDay <- function(date_val) {
        WdDate <- weekdays(as.Date(date_val, '%Y-%m-%d'))
        if  (!(WdDate == 'Saturday' || WdDate == 'Sunday')) {
                testData <- 'Weekday'
        } else {
                testData <- 'Weekend'
        }
        testData
}


````


Step 2:  Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

````{r,echo=FALSE}
MyData$day_type <- as.factor(sapply(MyData$date, WeekDay))
library(ggplot2)
StepsPerDay_imputed <- aggregate(steps ~ interval+day_type, MyData, mean)
myplot <- ggplot(StepsPerDay_imputed, aes(interval, steps)) +
        geom_line(stat = "identity", aes(colour = day_type)) +
        theme_gray() +
        facet_grid(day_type ~ ., scales="fixed", space="fixed") +
        labs(x="Interval", y=expression("No of Steps")) +
        ggtitle("Number Of Steps (Interval By Type Of Day ")
````


````{r}
print(myplot)
````

